**Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system**

### 2016-01-29 Schloss_PacBio16S_PeerJ_2016.Rmd
Use external error_summary file through pre-cluster analysis

regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
clrs <- rainbow(length(regions))
pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")
names(clrs) <- regions

library(json)
```

@@

@@

We established a simple curation procedure by culling any sequence that had a
string of the same base repeated more than 8 times  or did not start and end at
the expected alignment coordinates for that region of the 16S rRNA gene. This
reduced the experiment-wide error rate from `r format(init_error_rate, digits=3, nsmall=2)` to
`r format(basic_error_rate, digits=2, nsmall=2)`%. This basic procedure
resulted in the removal of between
`r format(min(100-basic_kept), digits=3, nsmall=1)`
(`r pretty_region[names(which.min(100-basic_kept))]`) and
`r format(max(100-basic_kept), digits=3, nsmall=1)`
(`r pretty_region[names(which.max(100-basic_kept))]`)% of the reads (Table 2). Although
the percentage of reads removed increased with the length of the fragment, there
was no obvious relationship between fragment length and error rate (Figure 2).

@@

#This is for the Figure 2D plot...
pred_overall <- aggregate(100 * mock_error_basic$error, by=list(round(mock_error_basic$pred_error, 3)), mean)

***Identifying correlates of increased sequencing error.***
In contrast to the 454 and Illumina-based platforms where the sequencing
quality decays with length, the consensus sequencing approach employed by the
PacBio sequencer is thought to generate a uniform distribution of errors. This
makes it impossible to simply trim sequences to high quality regions. Therefore,
we sought to identify characteristics within sequences that would allow us to
identify and remove those sequences with errors using three different
approaches. First, we hypothesized that errors in the barcode and primer would
be correlated with the error rate for the entire sequence. We observed a strong
relationship between the number of mismatches to the barcodes and primers and the
error rate of the rest of the sequence fragment (Figure 1B). Although
allowing no mismatches to the barcodes and primers yielded the lowest error rate,
that stringent criterion removed a large fraction of the reads from the dataset.
Allowing at most one mismatch only marginally increased the error rate while
retaining more sequences in the dataset (Figure 2). Second, we hypothesized that
increased sequencing coverage should yield lower error rates. We found that once
we had obtained 10-fold coverage of the fragments, the error rate did not change
appreciably (Figure 1C). When we compared the error rates of reads with at least
10-fold coverage to those with less coverage, we reduced the error rate by
`r paste(format(range(base_error_reduction), digits=3, nsmall=1), collapse=" to ")`%
(Figure 2). Third, based on the observed correlation between the
predicted and observed error rates, we sought to identify a minimum predicted
error rate that would allow us to reduce the observed error rate. The average
observed error rate for sequences with predicted error rates between 0.01 and
0.10% was linear (Figure 1D). We decided to use a threshold of 0.01% because a
large number of sequence reads were lost when we used a smaller threshold. When
we used this threshold, we were able to reduce the error rate by
`r paste(format(range(pred_error_reduction), digits=3, nsmall=1), collapse=" to ")`%
(Figure 2A). Finally, we quantified the effect of combining filters. We found that
any combination of filters that included the predicted error rate threshold had
the most significant impact on reducing the observed error rate. Furthermore,
the inclusion of the mismatch and coverage filters had a negligible impact on
error rates, but had a significant impact on the number of sequences included in
the analysis. For instance, among the V1-V9 data, requiring sequences to have a
predicted error rate less than 0.01% resulted in a `r format(base_error_reduction["pred_error_reduction","V1V9"], digits=3, nsmall=1)`% and resulted in the removal of `r format(base_nseqs_reduction["pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences. Adding the mismatch or coverage filter
had no effect on the reduction of error, but resulted in the removal of
`r format(base_nseqs_reduction["oligos_pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)` and
`r format(base_nseqs_reduction["coverage_pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences, respectively. Use of all filters had no impact on the reduction in the observed
error rate, but resulted in the removal of `r format(base_nseqs_reduction["all_filters_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences. The remainder of this paper only uses sequences with a predicted error rate less than 0.01%.

```{r precluster}
get_error_rate <- function(file_name){
	error_summary <- read.table(file=file_name, header=T, stringsAsFactors=FALSE)
	nochim <- error_summary[error_summary$numparents==1,]
    sum(nochim$weight * nochim$mismatches) / sum(nochim$weight * nochim$total)
}

regions <- c("V4", "V1V3", "V1V5", "V1V6", "V1V9", "V3V5")

unique_file_names <- paste0("data/mothur_pool/", regions, ".mock.screen.unique.good.filter.unique.error.summary")

unique_error <- sapply(unique_file_names, get_error_rate)

precluster_file_names <- paste0("data/mothur_pool/", regions,".mock.screen.unique.good.filter.unique.precluster.error.summary")

precluster_error <- sapply(precluster_file_names, get_error_rate)

pooled_error <- cbind(unique = unique_error, precluster = precluster_error)
rownames(pooled_error) <- regions

#          unique   precluster
#V4   0.003205151 0.0015805612
#V1V3 0.002529819 0.0006841053
#V1V5 0.001987322 0.0001938975
#V1V6 0.001910158 0.0003086235
#V1V9 0.001944654 0.0002743820
#V3V5 0.002076498 0.0003532300

reduction <- 100*(1-pooled_error[,"precluster"]/pooled_error[,"unique"])

min.region <- pretty_region[names(reduction[which.min(reduction)])]
max.region <- pretty_region[names(reduction[which.max(reduction)])]
```

***Pre-clustering sequences to further reduce sequencing noise.***
Previously, we implemented a pre-clustering algorithm where sequences were sorted
by their abundance in decreasing order and rare sequences are clustered with a
more abundant sequence if the rare sequences have fewer mismatches than a
defined threshold when compared to the more abundant sequence [@refs]. The recommended
threshold was a 1-nt difference per 100-nt of sequence data. For example, the
threshold for 250 bp fragment from the V4 region would be 2 nt or 14 for the
1458 bp V1-V9 fragments. This approach removes residual PCR and sequencing
errors while not overwhelming the resolution needed to identify OTUs that are
based on a 3% distance threshold. The tradeoff of this approach is that one would
unable to differentiate V1-V9 sequences that truly differed by less than 14 nt.
When we applied this approach to our PacBio data, we observed a reduction in the error rate
between `r min(reduction)` (`r min.region`) and `r max(reduction)`% (`r max.region`).
The final error rates varied between
`r round(100*min(pooled_error[,"precluster"]), digits=2)` (`r names(which.min(pooled_error[,"precluster"]))`) and `r round(100*max(pooled_error[,"precluster"]), digits=2)`% (`r names(which.max(pooled_error[,"precluster"]))`).
 The full-length (i.e. V1V9) fragments had an error rate of
`r round(100* pooled_error["V1V9","precluster"], digits=2)`%
(Figure 2B), this is similar to what we have previously observed using the 454 and
Illumina MiSeq platforms (0.02%){Schloss, 2011 #2466; Kozich, 2013 #2719}

>>>>

regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")

clrs <- rainbow(length(regions))
names(clrs) <- regions

library(rjson)
```

@@

error_summary <- read.table(file="data/process/error_summary.tsv", header=T)
raw_to_basic_lost <- 100 * (1 - error_summary[error_summary$method == "basic","nseqs"]/error_summary[error_summary$method == "raw","nseqs"])

@@

We established a simple curation procedure by culling any sequence that had a
string of the same base repeated more than 8 times  or did not start and end at
the expected alignment coordinates for that region of the 16S rRNA gene. This
reduced the experiment-wide error rate from `r format(mean(error_summary[error_summary$method=="raw","error"]), digits=3, nsmall=2)` to
`r format(mean(error_summary[error_summary$method=="basic","error"]), digits=2, nsmall=2)`%. This basic procedure resulted in the removal of between
`r format(min(raw_to_basic_lost), digits=3, nsmall=1)`
(`r pretty_region[names(which.min(raw_to_basic_lost))]`) and
`r format(max(raw_to_basic_lost), digits=3, nsmall=1)`
(`r pretty_region[names(which.max(raw_to_basic_lost))]`)% of the reads (Table 2). Although
the percentage of reads removed increased with the length of the fragment, there
was no obvious relationship between fragment length and error rate (Figure 2).

@@

#This is for the Figure 2D plot...
pred_overall <- aggregate(100 * mock_error_basic$error, by=list(round(mock_error_basic$pred_error, 3)), mean)
```

***Identifying correlates of increased sequencing error.***
In contrast to the 454 and Illumina-based platforms where the sequencing
quality decays with length, the consensus sequencing approach employed by the
PacBio sequencer is thought to generate a uniform distribution of errors. This
makes it impossible to simply trim sequences to high quality regions. Therefore,
we sought to identify characteristics within sequences that would allow us to
identify and remove those sequences with errors using three different
approaches. First, we hypothesized that errors in the barcode and primer would
be correlated with the error rate for the entire sequence. We observed a strong
relationship between the number of mismatches to the barcodes and primers and the
error rate of the rest of the sequence fragment (Figure 1B). Although
allowing no mismatches to the barcodes and primers yielded the lowest error rate,
that stringent criterion removed a large fraction of the reads from the dataset.
Allowing at most one mismatch only marginally increased the error rate while
retaining more sequences in the dataset (Figure 2). Second, we hypothesized that
increased sequencing coverage should yield lower error rates. We found that once
we had obtained 10-fold coverage of the fragments, the error rate did not change
appreciably (Figure 1C). When we compared the error rates of reads with at least
10-fold coverage to those with less coverage, we reduced the error rate by
`r paste(format(range(error_summary[error_summary$method == "coverage","error_red"]), digits=3, nsmall=1), collapse=" to ")`%
(Figure 2). Third, based on the observed correlation between the
predicted and observed error rates, we sought to identify a minimum predicted
error rate that would allow us to reduce the observed error rate. The average
observed error rate for sequences with predicted error rates between 0.01 and
0.10% was linear (Figure 1D). We decided to use a threshold of 0.01% because a
large number of sequence reads were lost when we used a smaller threshold. When
we used this threshold, we were able to reduce the error rate by
`r paste(format(range(error_summary[error_summary$method == "pred","error_red"]), digits=3, nsmall=1), collapse=" to ")`%
(Figure 2A). Finally, we quantified the effect of combining filters. We found that
any combination of filters that included the predicted error rate threshold had
the most significant impact on reducing the observed error rate. Furthermore,
the inclusion of the mismatch and coverage filters had a negligible impact on
error rates, but had a significant impact on the number of sequences included in
the analysis. For instance, among the V1-V9 data, requiring sequences to have a
predicted error rate less than 0.01% resulted in a
`r format(error_summary[error_summary$method == "pred" & error_summary$region == "V1V9","error_red"], digits=3, nsmall=1)`% reduction in error and resulted in the removal of
`r format(error_summary[error_summary$method == "pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`% of the sequences. Adding the mismatch or coverage filter
had no effect on the reduction of error, but resulted in the removal of
`r format(error_summary[error_summary$method == "pb_pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)` and
`r format(error_summary[error_summary$method == "coverage_pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`
% of the sequences, respectively. Use of all filters had no impact on the reduction in the observed
error rate, but resulted in the removal of `r format(error_summary[error_summary$method == "all" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`% of the sequences. The remainder of this paper only uses sequences with a predicted error rate less than 0.01%.

***Pre-clustering sequences to further reduce sequencing noise.***
Previously, we implemented a pre-clustering algorithm where sequences were sorted
by their abundance in decreasing order and rare sequences are clustered with a
more abundant sequence if the rare sequences have fewer mismatches than a
defined threshold when compared to the more abundant sequence [@refs]. The recommended
threshold was a 1-nt difference per 100-nt of sequence data. For example, the
threshold for 250 bp fragment from the V4 region would be 2 nt or 14 for the
1458 bp V1-V9 fragments. This approach removes residual PCR and sequencing
errors while not overwhelming the resolution needed to identify OTUs that are
based on a 3% distance threshold. The tradeoff of this approach is that one would
unable to differentiate V1-V9 sequences that truly differed by less than 14 nt.
When we applied this approach to our PacBio data, we observed a reduction in the error rate
between
`r format(min(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)`
(`r pretty_region[which.min(error_summary[error_summary$method == "precluster","nseqs_red"])]`)
and `r format(max(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)`%
(`r pretty_region[which.max(error_summary[error_summary$method == "precluster","nseqs_red"])]`).
The final error rates varied between
`r round(min(error_summary[error_summary$method == "precluster","error"]), digits=2)`
(`r pretty_region[which.min(error_summary[error_summary$method == "precluster","error"])]`)
and
`r round(max(error_summary[error_summary$method == "precluster","error"]), digits=2)`%
(`r pretty_region[which.max(error_summary[error_summary$method == "precluster","error"])]`).
The full-length (i.e. V1-V9) fragments had an error rate of
`r round(error_summary[error_summary$method == "precluster" & error_summary$region=="V1V9","error"], digits=2)`%
(Figure 2B), this is similar to what we have previously observed using the 454 and
Illumina MiSeq platforms (0.02%){Schloss, 2011 #2466; Kozich, 2013 #2719}
